Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-06 writeup/conStruct.tex

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each cluster, 
we define the covariance within cluster $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_cluster_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th cluster.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the cluster.
\gc{Our choice of a powered-exponential decay, as parameterized by the
  $\alpha$s, is a flexible, and standard choice in spatial statistics
  REF, and is not chosen to match  a particular population genetics model. 
The} $\phi^{(k)}$ is a parameter that describes the background covariance within the cluster. 
If two samples draw 100\% of their ancestry from cluster $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the cluster.
The ``shared drift'' parameter $\phi^{(k)}$ can be interpreted as the branch length 
connecting the $k$th population to the population ancestral to all modeled
layers \citep[see for example ][]{patterson_ancient_2012, peter_fstats}.

@@

\paragraph{Likelihood and inference}
If allele frequencies at each locus
were independent between loci and multivariate normally distributed across populations, 
then their allelic covariance $\widehat{\Omega}$ 
would be Wishart distributed with degrees of freedom equal to $L$, 
the number of loci genotyped.
We use this as a convenient approximation to the true distribution described above,
and so define the likelihood of the allelic covariance to be
\begin{equation}
P(\widehat{\Omega} \; | \; \Omega) = \mathcal{W} \left( L\widehat{\Omega} \; | \; \Omega,L\right) .
\end{equation}
Linkage disequilibrium (LD) between loci will decrease the effective number of degrees of freedom, 
a complication we discuss further below.

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;	| \; \widehat{\Omega}, L \right) \propto
P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors except for $\alpha_2$, which is uniform on $(0,2)$ and $w$, which is Dirichlet
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \citep{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \citep{R} called \texttt{conStruct} that functions as a wrapper 
around this inference machinery.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_cluster_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is strongly conceptually similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
Therefore, we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework.

\paragraph{Choice of cluster number and cross-validation}
There are a number of reasons, both statistical and biological, 
why there is no true (or right) number of layers for real datasets,
discussed further in the Discussion.
However, it is still important to assess whether additional spatial layers (larger $K$)
meaningfully model patterns in the data
or merely explain spurious variation introduced by noise
-- in other words, whether additional model complexity
provides significant explanatory power.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} results across different values of $K$ 
and between the spatial and nonspatial models.

@@

Cross-validation provides a valuable summary of how much explanatory power
is added by spatial structure within each layer, and each additional layer.
A simple $t$-test can be used to assess whether additional layers
show statistically significantly greater predictive accuracy.
However, we remind users that ``statistical significance does not imply real-world significance'',
and so small but statistically significant differences between models
should probably not be relied on too strongly.

@@

Similarly, we currently assume that all loci are independent, 
such that the number of loci in the dataset gives the degrees of 
freedom of the Wishart distribution in our likelihood calculations.  
Linkage disequilibrium (LD) between loci will decrease 
the effective number of degrees of freedom, 
artificially increasing users' confidence in their results.  
One possible solution is to estimate an \emph{effective} number of loci 
by introducing a parameter to modify the given degrees of freedom 
and thereby model linkage between loci \citep[e.g.][]{EEMS}.

@@

\section{Model rationale}
\subsection{Drift, admixture, and space}
\paragraph{Drift} 
We first provide a simple model of allele frequencies within a cluster. 
Imagine a sample $i$ that draws all of its ancestry from cluster $k$.
The allele frequency in sample $i$ at locus $\ell$, $F_{i,\ell}$ can be
written as the sum
\begin{equation}
F_{i,\ell} = \epsilon_{\ell} + \Delta^{(k)}_{\ell} +
\Delta^{(k,i)}_{\ell} + \Delta^{(i)}_{\ell}
\label{drift_terms_no_admix}
\end{equation}
The first term is the ancestral allele frequency $\epsilon_\ell$ shared by all samples; 
the second is the deviation from that ancestral frequency 
due to drift in the ancestral population of the $k$th cluster,
which is shared by all samples within the cluster. 
The third term is the deviation of the $i$th sample away from the $k$th cluster mean 
due to the spatial process of drift and migration within the cluster.
The final term is the deviation specific to the $i$th sample,
which captures drift not shared by all samples at the population level
(i.e., subpopulation-specific drift due to, e.g., inbreeding). 
We will assume that these four deviations are all uncorrelated with each other.

>>>>

To allow flexibility in the form of the decay of allelic covariance with geographic distance within each cluster, 
we define the covariance within cluster $k$ between samples $i$ and $j$ to be:
\begin{equation}
G^{(k)}_{i,j} 
    = 
    \alpha^{(k)}_0 \left( \text{exp} \left( -(\alpha^{(k)}_D D_{i,j}) ^ {\alpha^{(k)}_2}	\right) \right) + \phi^{(k)}
\label{within_cluster_covariance}
\end{equation}
where the superscript $(k)$ denotes parameters specific to the $k$th cluster.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
the $\alpha^{(k)}$ parameters control the shape of the decay of
covariance with distance in the cluster.
\gc{Our choice of a powered-exponential decay, as parameterized by the
  $\alpha$s, is a flexible, and standard choice in spatial statistics
  REF, and is not chosen to match  a particular population genetics model. 
The} $\phi^{(k)}$ is a parameter that describes the background covariance within the cluster. 
If two samples draw 100\% of their ancestry from cluster $k$, then their covariance under the model is $G^{(k)}_{i,j}$;
if they are furthermore geographically very close ($D_{i,j}=0$)
they will have covariance $\alpha^{(k)}_0 +  \phi^{(k)}$.
If the geographic distance between them is very large, 
their covariance will be equal to to the background level $\phi^{(k)}$ within the cluster.
The ``shared drift'' parameter $\phi^{(k)}$ can be interpreted as the branch length 
connecting the $k$th population to the population ancestral to all modeled
layers \citep[see for example ][]{patterson_ancient_2012,
  peter_fstats}. In the methods section \ref{rationale} we lay out a
simple model of allele frequencies underlying this covariance model.

@@

\paragraph{Likelihood and inference}
\gc{Assuming that the} allele frequencies at each locus
were independent between loci and multivariate normally distributed across populations, 
then their allelic covariance $\widehat{\Omega}$ 
would be Wishart distributed with degrees of freedom equal to $L$, 
the number of loci genotyped.
We use this as a convenient approximation to the true distribution described above,
and so define the likelihood of the allelic covariance to be
\begin{equation}
P(\widehat{\Omega} \; | \; \Omega) = \mathcal{W} \left( L\widehat{\Omega} \; | \; \Omega,L\right) .
\end{equation}
Linkage disequilibrium (LD) between loci will decrease the effective
number of degrees of freedom.  \gc{One possible solution, that we have
  not implemented yet, is to estimate an \emph{effective} number of loci 
by introducing a parameter to modify the given degrees of freedom 
and thereby informally model linkage between loci \citep[e.g.][]{EEMS}.}

We estimate the values of the parameters of the model using a Bayesian approach.
Acknowledging the dependence of the parametric covariance matrix $\Omega$ on its constituent parameters
$w,\alpha,\phi,\eta,\gamma$ and on the (observed) geographic distances $D$ 
with the notation $\Omega(w,\alpha,\phi,\eta,\gamma,D)$,
we denote the posterior probability density of the parameters as:
\begin{equation}
P\left( w,\alpha,\phi,\eta,\gamma \;	| \; \widehat{\Omega}, L \right) \propto
P\left(\widehat{\Omega} \; | \; \Omega(w,\alpha,\phi,\eta,\gamma,D) \right)
P(w)P(\alpha)P(\phi)P(\eta)P(\gamma) ,
\end{equation}
where $P(w)$, $P(\alpha)$, $P(\phi)$, $P(\eta)$, and $P(\gamma)$, are prior distributions.
All parameters are given (half-)Gaussian priors except for $\alpha_2$,
which is uniform on $(0,2)$ and $w$, which is an \gc{independent Dirichlet, of
  dimension $K$, for each individual}
(see Table \ref{tab:param_prior_tab} for specifics).
Parameters are independent between layers.
We use Hamiltonian Monte Carlo as implemented in STAN \citep{stan, NUTS, stan_lib, rstan} 
to estimate the posterior distribution on the parameters.
We also provide an R package \citep{R} called \texttt{conStruct} that functions as a wrapper 
around this inference machinery.

@@

A nice feature of our approach is that the model described in Eqn \eqref{cross_cluster_covariance} 
contains a nonspatial assignment model as a special case 
(see \secref{model_app} for a more in-depth discussion). 
By setting $\alpha_0^{(k)}$ to zero for all $k$, 
we obtain a nonspatial model in which each cluster has its own allele frequency at each SNP, 
and individuals draw a proportion of their ancestry from each cluster. 
This model is strongly conceptually similar to that of STRUCTURE \citep{STRUCTURE} 
and related models \citep[e.g.][]{ADMIXTURE}; 
the only difference is that our likelihood assumes that allele frequencies are normally distributed 
around their expectations, 
while the standard assignment methods assume that the error is binomially distributed \citep{Engelhardt2012}.  
Therefore, we can compare the fit of the different models 
--- spatial vs. nonspatial, across different values of $K$ --- 
by comparing their performance in a common framework. \gc{We also
  validate our claim that the results of our non-spatial method are a
  close match to those of STRUCTURE-based approaches.}

\paragraph{Choice of cluster number and cross-validation}
There are a number of \gc{reasons 
why there is no true (or right)} number of layers for real datasets,
discussed further in the Discussion.
However, it is still important to assess whether additional spatial layers (larger $K$)
meaningfully model patterns in the data
or merely explain spurious variation introduced by noise
-- in other words, whether additional model complexity
provides significant explanatory power.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} results across different values of $K$ 
and between the spatial and nonspatial models.

@@

Cross-validation provides a valuable summary of how much explanatory power
is added by spatial structure within each layer, and each additional layer.
A simple $t$-test can be used to assess whether additional layers
show statistically significantly greater predictive accuracy. \gc{DO
  WE DO t-test? IF NOT DROP THIS'}
However, we remind users that ``statistical significance does not imply real-world significance'',
and so small but statistically significant differences between models
should probably not be relied on too strongly.

@@

%%Similarly, we currently assume that all loci are independent, 
%such that the number of loci in the dataset gives the degrees of 
%freedom of the Wishart distribution in our likelihood calculations.  
%Linkage disequilibrium (LD) between loci will decrease 
%the effective number of degrees of freedom, 
%artificially increasing users' confidence in their results.

@@

\section{Model rationale}
\subsection{Drift, admixture, and space} \label{rationale}
\paragraph{Drift} 
We first provide a simple model of allele frequencies within a cluster. 
Imagine a sample $i$ that draws all of its ancestry from cluster $k$.
The allele frequency in sample $i$ at locus $\ell$, $F_{i,\ell}$ can be
written as the sum
\begin{equation}
F_{i,\ell} = \epsilon_{\ell} + \Delta^{(k)}_{\ell} +
\Delta^{(k,i)}_{\ell} + \Delta^{(i)}_{\ell}
\label{drift_terms_no_admix}
\end{equation}
The first term is the ancestral allele frequency $\epsilon_\ell$ shared by all samples; 
the second is the deviation from that ancestral frequency 
due to drift in the ancestral population of the $k$th cluster,
which is shared by all samples within the cluster. 
The third term is the deviation of the $i$th sample away from the $k$th cluster mean 
due to the spatial process of drift and migration within the cluster.
The final term is the deviation specific to the $i$th sample,
which captures drift not shared by all samples at the population level
(i.e., subpopulation-specific drift due to, e.g., inbreeding). 
We will assume that these four deviations are all uncorrelated with each other.
